\sysfull

### 2018-04-17 paper/paging.tex
done with paging

If the \texttt{working} buffer is full, a page fault on a memory access
requires \sys to swap out one of the pages in the volatile buffer (a 
\emph{victim page}) preserving updates made to that page, and to copy the
accessed page of protected data into the working buffer.
Algorithm~\ref{algo:pagefault} shows how \sys handles a page fault. A victim
page in the \texttt{working} buffer is selected for eviction according to
a \emph{first-in-first-out} replacement policy (Lines 2 and 12). If the page
being swapped out is not dirty and the page being accessed had not been dirtied
and swapped out to the \texttt{shadow} buffer since last boot, then handling a
page fault consists of copying the accessed page from
\texttt{\underline{store}} into the \texttt{working} buffer (Line 11).
Otherwise, the page fault handler must take the following extra steps.
%
If a task modified any byte in the victim page (i.e., using \texttt{WP}), then
the page is dirty. Before swapping in the accessed page, \sys must first
preserve the updated values in the dirty page so that they can be committed
back to their locations in memory when the dynamic task ends (but not earlier).
To preserve dirty values, \sys copies the victim page to the non-volatile
\texttt{\underline{shadow}} buffer (Line 5) and records the page's tag in a
non-volatile list (Lines 6-7) to support iteration over shadow pages having
traversing all slots in the buffer, necessary in the second phase of the
commit.
%
%Dirty pages remain in {\tt \underline{shadow}} buffer until the
%dynamic task completes, at which point \sys commits the modified
%pages from {\tt \underline{shadow}} back to \texttt{\underline{store}}.
%
If the accessed page was previously modified and swapped out since the last boot (Line 8), the most recent version of the page is in {\tt
\underline{shadow}}, not \texttt{\underline{store}} (Line 9).

@@

When the last task in a coalesced group completes, \sys must commit \emph{all} dirty pages in the \texttt{working} buffer and the swapped out pages in \texttt{\underline{shadow}} buffer back to their original locations in \texttt{\underline{store}}. If a task is allowed to (re-)execute after only \emph{some} pages have been committed to \texttt{\underline{store}}, that task might see protected variables in an inconsistent state.

To make the commit atomic, \sys commits the pages in two phases, implemented by Algorithm~\ref{algo:commit}.
%
The first phase copies dirty pages from the \texttt{working} buffer to the non-volatile \texttt{\underline{shadow}} buffer (Line 4). The second phase commits pages from \texttt{\underline{shadow}} to \texttt{\underline{store}} (Line 12).
If power fails during the first phase, the whole commit is aborted, and the dynamic coalesced task restarts from its first static task. If power fails in the second phase, the commit will be safely resumed on next boot.
The non-volatile metadata for the second phase are: the \texttt{\underline{committing}} bit, which is set before the first page is committed (Line 9) and cleared after the last page is committed (Line 16); \texttt{\underline{shadowCount}} counter, which is cleared once the phase completes (Line 14); \texttt{\underline{commitIndex}}, which indexes the next page to be committed (Line 11) and is cleared at the end of the phase (Line 15).

\sys non-volatile \texttt{\underline{shadowList/shadowCount/commitIndex}} implementation is a more efficient alternative to a simpler implementation that keeps a valid bit in each entry of \texttt{\underline{shadow}} and, on every boot into second phase, traverses all entries, committing the pages with a set bit, and clearing the bit. During the second commit phase, pages are not physically copied from \texttt{\underline{shadow}} to \texttt{\underline{store}}, but a table pointing at the right buffer for each page is updated. This is explained in more details in Section~\ref{sec:implementation}.

@@

One design choice \sys made was to insert dynamic check for paging on every
possible problematic memory operation using {\tt RP} and {\tt WP} macros.
Since dynamic checks incur overhead on every read and write, it has a higher
overhead for accessing memory than a static approach,
e.g., Alpaca, where a system statically analyze the memory operation and try to insert only optimal
number of memory protection code (usually at the beginning of each task).

However, \sys's dynamic paging has several benefits over static approach.
First, static approach has limitations, prohibiting certain types of programming pattern. For example,
in the presence of arbitrary pointer operation, function calls using function pointer, or an interrupt
within a task, the system cannot statically analyze the memory behavior. Thus, static
approaches either prohibits the programmer from using certain programming pattern~\cite{alpaca}, 
or will have to conservatively insert memory protection code that can be unnecessary.
Second, static approach cannot handle task coalescing, because with coalescing the scope to be
searched statically is not determined at compile time. 
By choosing to check every memory access dynamically, \sys supports programming pattern that
Alpaca does not support, such as arbitrary pointer operation, while also being able to
support task coalescing, which yields extra performance benefits.
%
%Moreover, if the programmer wants to optimize for high performance, it is possible for the
%programmer to place {\tt RP} and {\tt WP} only minimally by carefully examining the code.
%We argue that the

>>>>

If the \texttt{working} buffer is full, a page fault on a memory access
requires \sys to swap out one of the pages in the volatile buffer (a 
\emph{victim page}) preserving updates made to that page, and to copy the
accessed page of protected data into the working buffer.
Algorithm~\ref{algo:pagefault} shows how \sys handles a page fault. A victim
page in the \texttt{working} buffer is selected for eviction according to
a \emph{first-in-first-out} (FIFO) replacement policy (Lines 2 and 12). 
%
While we opted for the simplicity of FIFO, any other replacement policy, such
as Least Recently Used (LRU) could work in its place.
%
If a task modified any byte in the victim page (i.e., using \texttt{WP}), then
the page is dirty.  If the page being swapped out is not dirty and the page
being accessed had not been dirtied and swapped out to the \texttt{shadow}
buffer since the last reboot, then handling a page fault requires copying the
accessed page from its home in the  \texttt{\underline{store}} into the
\texttt{working} buffer (Line 11).  Otherwise, the page fault handler must take
the following extra steps.
%
Before swapping in the accessed page, \sys must first preserve the updated
values in the dirty page so that they can be committed back to their locations
in memory when the task commits (but not earlier).  To preserve the dirty
values, \sys copies the victim page to the non-volatile
\texttt{\underline{shadow}} buffer (Line 5) and records the page's tag in a
non-volatile commit list (Lines 6-7). \sys uses the commit list to iterate over shadow pages 
at commit time, to copy the contents of shadow pages back to the main memory {\tt store}. 
%
%Dirty pages remain in {\tt \underline{shadow}} buffer until the
%dynamic task completes, at which point \sys commits the modified
%pages from {\tt \underline{shadow}} back to \texttt{\underline{store}}.
%
If the accessed page was previously modified and swapped out since the last
power failure (Line 8), the most recent version of the page is in {\tt
\underline{shadow}}, not \texttt{\underline{store}} (Line 9).

@@

When the last task in a sequence of coalesced tasks completes, \sys must commit
\emph{all} dirty pages in the \texttt{working} buffer and in the
\texttt{\underline{shadow}} buffer, copying them back to their original
locations in the main memory \texttt{\underline{store}}. If a task were allowed
to (re-)execute after only some pages have been committed to
\texttt{\underline{store}}, the task might see protected variables in an
inconsistent state; \sys's two-phase commit prevents this inconsistency.

To make the commit atomic, \sys commits the pages in two phases, implemented by
Algorithm~\ref{algo:commit}.
%
The first phase copies dirty pages from the \texttt{working} buffer to the
non-volatile \texttt{\underline{shadow}} buffer (Line 4). The second phase
commits pages from \texttt{\underline{shadow}} to \texttt{\underline{store}}
(Line 12).  If power fails during the first phase, the whole commit is aborted,
and the execution restarts from the most recently committed point (i.e., from
the first task in a coalesced sequence of tasks). If power fails in the second
phase, the commit process safely resumes on reboot.  The second phase of commit
depends on some run-time metadata. The \texttt{\underline{committing}} bit
indicates that a commit is in progress and is set before the first page is
committed (Line 9) and cleared after the last page is committed (Line 16).  The
\texttt{\underline{shadowCount}} records the number of dirty shadow pages to
commit and \sys clears the counter when commit completes (Line 14).  The
\texttt{\underline{commitIndex}} indexes the next page to be committed (Line
11) and \sys clears the index at the end of the phase (Line 15).
%
\sys's commit is efficient because it maintains an index dirty pages instead of
iterating through all potentially dirty pages to check a dirty bit during
commit. 
%
\sys's implementation is also efficient because the memory manager does not
copy from \texttt{\underline{shadow}} to \texttt{\underline{store}}, instead
swapping pointers in an indirection table that maintains the pages as a double
buffer.  Section~\ref{sec:implementation} describes double buffering in more
detail.

@@

\sys asks the programmer to use its {\tt RP} and {\tt WP} paging API methods on
every access to a protected variable.  These API invocations present a risk of
high overhead because there is a dynamic check on every read and write. An
alternative approach explored by prior work~\cite{alpaca} relied on a compiler
to analyze memory operations and try to insert memory
protection code mostly at the beginning of a task, leaving most memory accesses uninstrumented.

Despite the risk of per-access overhead, \sys's dynamic memory protection
scheme brings several benefits over a static approach.  First, the limitations
of static analysis preclude some uses of pointers due to potential pointer
aliasing. For example, in the presence of arbitrary pointer operation, a
function call using a function pointer, or an interrupt within a task, the
system cannot statically analyze the memory behavior.   Second, a static approach
cannot handle task coalescing because a protected variable's lifetime (i.e., from first use to commit) is unknown 
at compile time. \sys's dynamic, per-access instrumentation supports arbitrary use of pointers 
and enables task coalescing.
%
%Moreover, if the programmer wants to optimize for high performance, it is possible for the
%programmer to place {\tt RP} and {\tt WP} only minimally by carefully examining the code.
%We argue that the
